Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-05 paper/zhang-cook-laa-langrene-menendez.Rmd
minor fixes: keywords,interestingness, optim prob and captions

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu

abstract: >
  Guided tour searches for interesting low-dimensional views of high-dimensional data via optimising a projection pursuit index function. The first paper of projection pursuit by @friedman1974projection stated that "the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." While much work has been done in proposing  indices in the literature, less has been done on evaluating the performance of the optimisers. In this paper, we implement a data collection object in the optimisation of projeciton pursuit guided tour and introduce visual diagnostics based on the data object collected. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, \pkg{ferrn}, has been created to implement the diagnostics.
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

<!-- Notation is first established before defining projection pursuit. -->
Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. A number of index functions have been proposed in the literature to detect different data structures [@friedman1974projection; @hall1989polynomial; @cook1993projection; @posse1995projection] and to classify groups [@lee2005projection; @lee2010projection].

@@

```{r tour-path, out.height="20%",out.width="50%", fig.cap="Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation."}
include_graphics(here::here("img/tour-path.png"))
```

@@

In projection pursuit the optimisation problem aims at finding the global optimum that gives the most interesting projection according to an index function. That is, it starts with a given randomly selected basis $\mathbf{A}_1$ and aims at finding an optimal final projection basis $\mathbf{A}_T$ in each iteration that satisfies the following optimisation problem:

\begin{align}
&\arg \max_{\mathbf{A} \in \mathcal{A}} f(\mathbf{X} \cdot \mathbf{A}) \\
&s.t.  \mathbf{A}^{\prime} \mathbf{A} = I_d
\end{align}

@@

```{r toy-search, fig.cap="A comparison of search by two optimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \\code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iterations because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8."}
```

@@

```{r toy-interp, fig.cap = "Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotonic."}
```

>>>>

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu 
abstract: |
  Guided tour searches for interesting low-dimensional views of high-dimensional data via optimising a projection pursuit index function. The first paper of projection pursuit by @friedman1974projection stated that "the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." While much work has been done in proposing  indices in the literature, less has been done on evaluating the performance of the optimisers. In this paper, we implement a data collection object in the optimisation of projeciton pursuit guided tour and introduce visual diagnostics based on the data object collected. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, \pkg{ferrn}, has been created to implement the diagnostics.

  \noindent \textit{keywords}: graphics, multivariate, optimization
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. In the context of projection pursuit, interestingness means departure from normality [@diaconis1984asymptotics; @huber1985projection]. \textcolor{red}{only include a couple of indices, using e.g.}A number of index functions have been proposed in the literature to detect different data structures [@friedman1974projection; @hall1989polynomial; @cook1993projection; @posse1995projection] and to classify groups [@lee2005projection; @lee2010projection].

@@

```{r tour-path, out.height="20%",out.width="50%", fig.cap="An illustration for demonstrating the steps in a tour path. Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation."}
include_graphics(here::here("img/tour-path.png"))
```

@@

In projection pursuit the optimisation problem aims at finding the global and local maxima that give interesting projections according to an index function. That is, it starts with a given randomly selected basis $\mathbf{A}_1$ and aims at finding an optimal final projection basis $\mathbf{A}_T$ in each iteration that satisfies the following optimisation problem:

\begin{align}
\arg \max_{\mathbf{A} \in \mathcal{A}} f(\mathbf{X} \cdot \mathbf{A})  ~~~ s.t. ~~~ \mathbf{A}^{\prime} \mathbf{A} = I_d
\end{align}

@@

```{r toy-search, fig.cap="A comparison of the searches by two optimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \\code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iterations because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8."}
```

@@

```{r toy-interp, fig.cap = "Examing the index values as the optimisation progress for two optimisers using trace plot. The lines indicate the interpolation and the dots indicate a new target is generated."}
```

<!-- The right hand side has the jump out, which has a lower index value - also explain this in the text. -->

<!-- Both traces are smooth while the change of index value in each iteration may not be monotonic. -->

### 2021-02-05 paper/zhang-cook-laa-langrene-menendez.tex
minor fixes: keywords,interestingness, optim prob and captions

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projeciton
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and workflows can be applied to a
broad class of optimisers, to assess their performance. An R package,
\pkg{ferrn}, has been created to implement the diagnostics.
}

@@

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. A number of index functions have been proposed in
the literature to detect different data structures
\citep{friedman1974projection, hall1989polynomial, cook1993projection, posse1995projection}
and to classify groups \citep{lee2005projection, lee2010projection}.

@@

\caption[Each square (frame) represents the projected data with a corresponding basis]{Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.}\label{fig:tour-path}
\end{figure}
\end{Schunk}

@@

In projection pursuit the optimisation problem aims at finding the
global optimum that gives the most interesting projection according to
an index function. That is, it starts with a given randomly selected
basis \(\mathbf{A}_1\) and aims at finding an optimal final projection
basis \(\mathbf{A}_T\) in each iteration that satisfies the following
optimisation problem:

\begin{align}
&\arg \max_{\mathbf{A} \in \mathcal{A}} f(\mathbf{X} \cdot \mathbf{A}) \\
&s.t.  \mathbf{A}^{\prime} \mathbf{A} = I_d
\end{align}

@@

\caption[A comparison of search by two optimisers]{A comparison of search by two optimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iterations because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8.}\label{fig:toy-search}
\end{figure}
\end{Schunk}

@@

\caption[Trace plots of the interpolating bases with the same optimisation routine as the previous figure]{Trace plots of the interpolating bases with the same optimisation routine as the previous figure. Both traces are smooth while the change of index value in each iteration may not be monotonic.}\label{fig:toy-interp}
\end{figure}
\end{Schunk}

@@

\caption[The PCA plot of 1D projection problem on the 5-variable dataset, \code{boa5}, using the holes index with two optimisers]{The PCA plot of 1D projection problem on the 5-variable dataset, \code{boa5}, using the holes index with two optimisers: SA and PD. All the bases in PD has been flipped and a grey dashed line has been annotated to indicate the symmetry of two starting bases.}\label{fig:toy-pca}
\end{figure}
\end{Schunk}

@@

\caption{Six selected frames from the animated PCA plots in the previous figure. With animation, the start and end position of each optimisation is easier to identify.  A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

>>>>

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projeciton
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and workflows can be applied to a
broad class of optimisers, to assess their performance. An R package,
\pkg{ferrn}, has been created to implement the diagnostics.

\noindent \textit{keywords}: graphics, multivariate, optimization
}

@@

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. In the context of projection pursuit,
interestingness means departure from normality
\citep{diaconis1984asymptotics, huber1985projection}.
\textcolor{red}{only include a couple of indices, using e.g.}A number of
index functions have been proposed in the literature to detect different
data structures
\citep{friedman1974projection, hall1989polynomial, cook1993projection, posse1995projection}
and to classify groups \citep{lee2005projection, lee2010projection}.

@@

\caption[An illustration for demonstrating the steps in a tour path]{An illustration for demonstrating the steps in a tour path. Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.}(\#fig:tour-path)
\end{figure}
\end{Schunk}

@@

In projection pursuit the optimisation problem aims at finding the
global and local maxima that give interesting projections according to
an index function. That is, it starts with a given randomly selected
basis \(\mathbf{A}_1\) and aims at finding an optimal final projection
basis \(\mathbf{A}_T\) in each iteration that satisfies the following
optimisation problem:

\begin{align}
\arg \max_{\mathbf{A} \in \mathcal{A}} f(\mathbf{X} \cdot \mathbf{A})  ~~~ s.t. ~~~ \mathbf{A}^{\prime} \mathbf{A} = I_d
\end{align}

@@

\caption[A comparison of the searches by two optimisers]{A comparison of the searches by two optimisers: SA (left) and SAJO (right) on a 2D projection of a six-variable dataset, \code{boa6}, using the holes index. Both optimisers find the final basis with similar index value while it takes SAJO more iterations because the algorithm allows a probabilistic acceptance of bases with lower index value, as observed in iteration 4 and 6-8.}(\#fig:toy-search)
\end{figure}
\end{Schunk}

@@

\caption[Examing the index values as the optimisation progress for two optimisers using trace plot]{Examing the index values as the optimisation progress for two optimisers using trace plot. The lines indicate the interpolation and the dots indicate a new target is generated.}(\#fig:toy-interp)
\end{figure}
\end{Schunk}

@@

\caption[The PCA plot of 1D projection problem on the 5-variable dataset, \code{boa5}, using the holes index with two optimisers]{The PCA plot of 1D projection problem on the 5-variable dataset, \code{boa5}, using the holes index with two optimisers: SA and PD. All the bases in PD has been flipped and a grey dashed line has been annotated to indicate the symmetry of two starting bases.}(\#fig:toy-pca)
\end{figure}
\end{Schunk}

@@

\caption{Six selected frames from the animated PCA plots in the previous figure. With animation, the start and end position of each optimisation is easier to identify.  A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}}(\#fig:toy-pca-animated)
\end{figure}
\end{Schunk}
